Binning Strategies and Related Loss for Large Data

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Update Baseball data to 2014

# connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
#                      port=3306, user="dbaccess", dbname="baseball")
# pitch <- new("dataDB", co=connect, table="Pitching")
# 
# d1 <- dbData(pitch, vars=c( "G", "SO")) # Games, Strike Outs
# 
# xtd.data <- function(x) {
# # replicate each line as often as given by the last entry
#   k <- dim(x)[2]
# 
#   res <- data.frame(sapply(1:(k-1), function(var) {
#   		return(rep(x[,var], x[,k]))
# 		}
# 	))
# 	names(res) <- names(x)[1:(k-1)]
# 	return(res)
# }
# d1.ext <- xtd.data(d1)
# dbDisconnect(connect)
# pitchdat <- d1.ext
# save(d1, d1.ext,pitchdat, file="./data/pitching.Rdata")
load("./data/pitching.Rdata")
@

<<scatterplots, echo=F,include=F>>=
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
  ggsave("./figure/OverplottingAlpha.png", width=5, height=5, dpi=900)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
  ggsave("./figure/Overplotting.png", width=5, height=5, dpi=900)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
  ggsave("./figure/OverplottingCircles.png", width=5, height=5, dpi=900)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.png", width=5, height=5, dpi=900)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <-  d1.hex.obj@count

@@

In the case of modestly sized data, scatterplots are great tools for showing bivariate data relationships. With large data, scatterplots suffer from over-plotting of points, which masks relevant structure. Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games played in a season is plotted against the number of strikeouts a pitcher threw over the course of a season. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.

>>>>

# # Downloaded from http://seanlahman.com/baseball-archive/statistics 2014 update
# pitchdat <- read.csv("./data/Pitching2014.csv")[,c("G","SO")]
# source("RectangularBinningFunctions.R")
# d1 <- RectBin2d(xs=pitch2014$G,ys=pitch2014$SO, 
#                      originx = .5, originy = -.5, 
#                      widthx = 1, widthy = 1, type="standard")
# names(d1) <- c( "G","SO","Freq")
# save(d1, pitchdat, file="./data/pitching.Rdata")

load("./data/pitching.Rdata")
@

<<scatterplots, echo=F,include=F>>=
qplot(data=pitchdat, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
  ggsave("./figure/OverplottingAlpha.png", width=5, height=5, dpi=900)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
  ggsave("./figure/Overplotting.png", width=5, height=5, dpi=900)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
  ggsave("./figure/OverplottingCircles.png", width=5, height=5, dpi=900)
qplot(data=pitchdat, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.png", width=5, height=5, dpi=900)
d1.hex.obj <- hexbin(pitchdat$G, pitchdat$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <-  d1.hex.obj@count

@@

In the case of modestly sized data, scatterplots are great tools for showing bivariate data relationships. With large data, scatterplots suffer from over-plotting of points, which masks relevant structure. Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 144 seasons (from the years of 1871 -- 2014) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games played in a season is plotted against the number of strikeouts a pitcher threw over the course of a season. While the data set is only medium sized with \Sexpr{dim(pitchdat)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.
